---
title: "vivainsights Download/Usage Report"
author: "Martin Chan"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
---

The notebook collects, aggregates, and visualizes the download numbers for the 'vivainsights' R and Python packages:

- [R](https://microsoft.github.io/vivainsights/)
- [Python](https://microsoft.github.io/vivainsights-py/)

The official repositories for open-source distribution of R and Python packages are the Comprehensive R Archive Network (CRAN) and Python Package Index (PyPI) respectively. 

There is no direct way to track package _usage_, but the figures can be estimated by combining documentation views and other data sources.

```{r setup, echo=FALSE,message=FALSE,warning=FALSE}
knitr::opts_chunk$set(
	echo = FALSE,
	message = FALSE,
	warning = FALSE
)

library(dplyr)
library(purrr)
library(tidyr)
library(ggplot2)
library(shiny)
library(readr)
library(cranlogs)
library(gh) # GitHub API
library(httr)
library(jsonlite)
library(summaryBox) # remotes::install_github("deepanshu88/summaryBox")

pkg_name <- "vivainsights"
```


```{r echo=FALSE,message=FALSE,warning=FALSE}
start_date <- as.Date("2023-08-01")
end_date <- as.Date(Sys.Date(), format = "%Y-%m-%d")
cran_dl <-
  cranlogs::cran_downloads(packages = pkg_name,
                           from = start_date,
                           to = end_date)

cran_dl <-
  cran_dl %>%
  dplyr::select(date, CRAN = "count")
```

```{r}
api_endpoint <- paste0("https://pypistats.org/api/packages/", pkg_name, "/overall")

response <- httr::GET(api_endpoint)

if (status_code(response) == 200) {
  data <- jsonlite::fromJSON(content(response, "text"))
  pypi_dl <- data$data
  
} else {
  
  cat(paste0("Error: ", status_code(response), " - ", status_reason(response)))
  
}
```


```{r join-data, echo=FALSE,message=FALSE,warning=FALSE}
# pypistats returns one row per date per category ("with_mirrors" and
# "without_mirrors"). Use with_mirrors only to avoid double-counting; this is
# the convention pypistats.org itself displays.
pypi_dl_agg <-
  pypi_dl %>%
  dplyr::filter(category == "with_mirrors") %>%
  group_by(date) %>%
  summarise(PyPI = sum(downloads, na.rm = TRUE)) %>%
  mutate(date = as.Date(date, format = "%Y-%m-%d"))

full_dl_snapshot <-
  cran_dl %>%
  full_join(
    pypi_dl_agg,
    by = "date"
  )

full_dl_snapshot %>%
  write_rds(file = paste0(
    "data/",
    pkg_name,
    "_",
    min(full_dl_snapshot$date, na.rm = TRUE),
    "_",
    max(full_dl_snapshot$date, na.rm = TRUE),
    ".rds"
  ))

# Read RDS back in. When the same date appears in multiple snapshots, prefer
# the value from the most recently-written snapshot (filenames embed the
# snapshot's end date). This way the mirror-double-count fix applied to new
# snapshots takes precedence over older, inflated PyPI values.
snapshot_files <-
  list.files(path = here::here("data"), full.names = TRUE) %>%
  .[grepl(x = ., pattern = "[.]rds$", ignore.case = TRUE)]

full_dl <-
  snapshot_files[order(file.mtime(snapshot_files))] %>%
  purrr::imap(function(path, i) {
    read_rds(path) %>% mutate(.snapshot_rank = i)
  }) %>%
  bind_rows() %>%
  arrange(date, .snapshot_rank) %>%
  group_by(date) %>%
  summarise(
    CRAN = dplyr::last(CRAN[!is.na(CRAN)], default = NA_real_),
    PyPI = dplyr::last(PyPI[!is.na(PyPI)], default = NA_real_),
    .groups = "drop"
  ) %>%
  arrange(date)
```

```{r lifetime-totals, echo=FALSE,message=FALSE,warning=FALSE}
# CRAN lifetime: query cranlogs for the full available history.
cran_lifetime <- tryCatch({
  cranlogs::cran_downloads(
    packages = pkg_name,
    from = "2013-01-01",
    to = as.character(Sys.Date() - 1)
  ) %>%
    summarise(total = sum(count, na.rm = TRUE)) %>%
    pull(total)
}, error = function(e) NA_integer_)

# PyPI lifetime: pepy.tech exposes a true lifetime total. Requires PEPY_API_KEY
# in the environment (set as a GitHub Actions secret). Falls back to summing the
# accumulated daily snapshots when the key is unavailable.
pepy_key <- Sys.getenv("PEPY_API_KEY", unset = "")
pypi_lifetime <- NA_integer_
pypi_lifetime_source <- "snapshots"

if (nzchar(pepy_key)) {
  pepy_resp <- tryCatch(
    httr::GET(
      paste0("https://api.pepy.tech/api/v2/projects/", pkg_name),
      httr::add_headers(`X-API-Key` = pepy_key)
    ),
    error = function(e) NULL
  )
  if (!is.null(pepy_resp) && httr::status_code(pepy_resp) == 200) {
    pepy_data <- jsonlite::fromJSON(httr::content(pepy_resp, "text", encoding = "UTF-8"))
    if (!is.null(pepy_data$total_downloads)) {
      pypi_lifetime <- as.numeric(pepy_data$total_downloads)
      pypi_lifetime_source <- "pepy.tech"
    }
  }
}

if (is.na(pypi_lifetime)) {
  pypi_lifetime <- sum(full_dl$PyPI, na.rm = TRUE)
}
```

This report captures `r round(n_distinct(full_dl$date))` days of data ranging from `r min(full_dl$date)` to `r max(full_dl$date)`.

# Summary

The boxes below show the headline summary statistics of the R and Python repositories. 

Numbers are aggregated for the last 30 days, excluding the current day and the day before due to incomplete data. 

```{r echo=FALSE,message=FALSE,warning=FALSE}
full_dl_l30 <-
  full_dl %>%
  arrange(date) %>%
  head(nrow(.) - 2) %>% # Remove last one day
  tail(30) %>% # last 30 days
  summarise(
    across(
      .cols = -date,
      .fns = ~sum(., na.rm = TRUE)
    )
  )

shiny::fluidRow(
  summaryBox(
    "CRAN Downloads (Last 30 days)",
    full_dl_l30[["CRAN"]],
    width = 3,
    icon = "fa fa-cloud-download",
    style = "info"
  ),
  
  summaryBox(
    "PyPI Downloads (Last 30 days)",
    full_dl_l30[["PyPI"]],
    width = 3,
    icon = "fa fa-cloud-download",
    style = "info"
  ),

  summaryBox(
    "CRAN Downloads (Lifetime)",
    cran_lifetime,
    width = 3,
    icon = "fa fa-trophy",
    style = "success"
  ),

  summaryBox(
    paste0("PyPI Downloads (Lifetime, ", pypi_lifetime_source, ")"),
    pypi_lifetime,
    width = 3,
    icon = "fa fa-trophy",
    style = "success"
  )
)

```


# Download statistics - CRAN, PyPI

The following plot shows the download statistics of the libraries from CRAN (R) and PyPI (Python). 

In general, download statistics from CRAN and PyPI should be more representative of actual user download and updates.

## Plots

### Daily downloads from CRAN and PyPI

```{r echo=FALSE,message=FALSE,warning=FALSE}
ts_plot <- 
  full_dl %>%
  select(date, CRAN, PyPI) %>%
  ggplot(aes(x = date)) +
  geom_smooth(aes(y = CRAN, color = "CRAN"), size = 1) +
  geom_smooth(aes(y = PyPI, color = "PyPI"), size = 1)+
  scale_color_manual(values = c("CRAN" = "#1f77b4", "PyPI" = "#ff7f0e")) +
  labs(title = "CRAN and PyPI Downloads",
       x = "Date",
       y = "Downloads",
       color = "Source") +
  theme_minimal() +
  theme(plot.title = element_text(size = 16, face = "bold"),
        axis.title = element_text(size = 14),
        axis.text = element_text(size = 12),
        legend.title = element_text(size = 12),
        legend.text = element_text(size = 12))

plotly::ggplotly(ts_plot)
```

## Notes

- The download figures may sometimes show irregularities, as new releases tend to trigger more downloads due to automated package archiving.
- Package updates also count towards downloads