---
title: "`r params$pkg_name` Download/Usage Report"
author: "Martin Chan"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
params:
  pkg_name: "vivainsights"
  start_date: "2023-08-01"     # First public release of the vivainsights R/Python packages
  keep_snapshots: 26            # Retain ~1 year of bi-monthly snapshots in data/
---

This notebook collects, aggregates, and visualizes download numbers for the **`r params$pkg_name`** R and Python packages:

- [R](https://microsoft.github.io/vivainsights/) — distributed on the Comprehensive R Archive Network (CRAN).
- [Python](https://microsoft.github.io/vivainsights-py/) — distributed on the Python Package Index (PyPI).

Download counts are an imperfect proxy for usage — CI runs, mirrors, and automated archiving all inflate them — but they are the most consistently available signal across registries. See the [methodology section in the README](./README.md#methodology) for the caveats.

```{r setup, echo=FALSE,message=FALSE,warning=FALSE}
knitr::opts_chunk$set(
	echo = FALSE,
	message = FALSE,
	warning = FALSE
)

library(dplyr)
library(purrr)
library(ggplot2)
library(readr)
library(cranlogs)
library(httr)
library(jsonlite)
library(htmltools)
library(summaryBox) # remotes::install_github("deepanshu88/summaryBox")

pkg_name <- params$pkg_name
```


```{r cran-fetch, echo=FALSE,message=FALSE,warning=FALSE}
start_date <- as.Date(params$start_date)
end_date   <- Sys.Date()

cran_dl <-
  cranlogs::cran_downloads(packages = pkg_name,
                           from = start_date,
                           to = end_date) %>%
  dplyr::select(date, CRAN = "count")
```

```{r pypi-fetch, echo=FALSE,message=FALSE,warning=FALSE}
api_endpoint <- paste0("https://pypistats.org/api/packages/", pkg_name, "/overall")

response <- httr::GET(api_endpoint)

if (httr::status_code(response) == 200) {
  pypi_payload <- jsonlite::fromJSON(httr::content(response, "text", encoding = "UTF-8"))
  pypi_dl <- pypi_payload$data
} else {
  warning(sprintf(
    "pypistats.org returned %s (%s); falling back to an empty PyPI frame for this render.",
    httr::status_code(response),
    httr::http_status(response)$reason
  ))
  pypi_dl <- data.frame(
    category = character(),
    date = character(),
    downloads = integer(),
    stringsAsFactors = FALSE
  )
}
```


```{r join-data, echo=FALSE,message=FALSE,warning=FALSE}
# pypistats returns one row per date per category ("with_mirrors" and
# "without_mirrors"). Use with_mirrors only to avoid double-counting; this is
# the convention pypistats.org itself displays.
pypi_dl_agg <-
  pypi_dl %>%
  dplyr::filter(category == "with_mirrors") %>%
  group_by(date) %>%
  summarise(PyPI = sum(downloads, na.rm = TRUE)) %>%
  mutate(date = as.Date(date, format = "%Y-%m-%d"))

full_dl_snapshot <-
  cran_dl %>%
  full_join(
    pypi_dl_agg,
    by = "date"
  )

full_dl_snapshot %>%
  write_rds(file = paste0(
    "data/",
    pkg_name,
    "_",
    min(full_dl_snapshot$date, na.rm = TRUE),
    "_",
    max(full_dl_snapshot$date, na.rm = TRUE),
    ".rds"
  ))

# Prune older snapshots so data/ stays bounded. Keeps the N most-recently
# modified rds files for this package (configurable via params$keep_snapshots).
snapshot_pattern <- paste0("^", pkg_name, "_.*\\.rds$")
all_pkg_snapshots <- list.files(
  path = here::here("data"),
  pattern = snapshot_pattern,
  full.names = TRUE
)
if (length(all_pkg_snapshots) > params$keep_snapshots) {
  to_remove <- all_pkg_snapshots %>%
    .[order(file.mtime(.), decreasing = TRUE)] %>%
    .[(params$keep_snapshots + 1L):length(.)]
  file.remove(to_remove)
}

# Read RDS back in. When the same date appears in multiple snapshots, prefer
# the value from the most recently-written snapshot (filenames embed the
# snapshot's end date). This way the mirror-double-count fix applied to new
# snapshots takes precedence over older, inflated PyPI values.
snapshot_files <-
  list.files(path = here::here("data"), full.names = TRUE) %>%
  .[grepl(x = ., pattern = "[.]rds$", ignore.case = TRUE)]

full_dl <-
  snapshot_files[order(file.mtime(snapshot_files))] %>%
  purrr::imap(function(path, i) {
    read_rds(path) %>% mutate(.snapshot_rank = i)
  }) %>%
  bind_rows() %>%
  arrange(date, .snapshot_rank) %>%
  group_by(date) %>%
  summarise(
    CRAN = dplyr::last(CRAN[!is.na(CRAN)], default = NA_real_),
    PyPI = dplyr::last(PyPI[!is.na(PyPI)], default = NA_real_),
    .groups = "drop"
  ) %>%
  arrange(date)
```

```{r lifetime-totals, echo=FALSE,message=FALSE,warning=FALSE}
# CRAN lifetime: query cranlogs for the full available history.
cran_lifetime <- tryCatch({
  cranlogs::cran_downloads(
    packages = pkg_name,
    from = "2013-01-01",
    to = as.character(Sys.Date() - 1)
  ) %>%
    summarise(total = sum(count, na.rm = TRUE)) %>%
    pull(total)
}, error = function(e) NA_integer_)

# PyPI lifetime: pepy.tech exposes a true lifetime total. Requires PEPY_API_KEY
# in the environment (set as a GitHub Actions secret). Falls back to summing the
# accumulated daily snapshots when the key is unavailable.
pepy_key <- Sys.getenv("PEPY_API_KEY", unset = "")
pypi_lifetime <- NA_integer_
pypi_lifetime_source <- "snapshots"

if (nzchar(pepy_key)) {
  pepy_resp <- tryCatch(
    httr::GET(
      paste0("https://api.pepy.tech/api/v2/projects/", pkg_name),
      httr::add_headers(`X-API-Key` = pepy_key)
    ),
    error = function(e) NULL
  )
  if (!is.null(pepy_resp) && httr::status_code(pepy_resp) == 200) {
    pepy_data <- jsonlite::fromJSON(httr::content(pepy_resp, "text", encoding = "UTF-8"))
    if (!is.null(pepy_data$total_downloads)) {
      pypi_lifetime <- as.numeric(pepy_data$total_downloads)
      pypi_lifetime_source <- "pepy.tech"
    }
  }
}

if (is.na(pypi_lifetime)) {
  pypi_lifetime <- sum(full_dl$PyPI, na.rm = TRUE)
}
```

This report captures `r round(n_distinct(full_dl$date))` days of data ranging from `r min(full_dl$date)` to `r max(full_dl$date)`.

# Summary

The boxes below show the headline summary statistics of the R and Python repositories. 

Numbers are aggregated for the last 30 days, excluding the current day and the day before due to incomplete data. 

```{r echo=FALSE,message=FALSE,warning=FALSE}
full_dl_l30 <-
  full_dl %>%
  arrange(date) %>%
  head(nrow(.) - 2) %>% # Remove last one day
  tail(30) %>% # last 30 days
  summarise(
    across(
      .cols = -date,
      .fns = ~sum(., na.rm = TRUE)
    )
  )

htmltools::div(
  class = "row",
  summaryBox(
    "CRAN Downloads (Last 30 days)",
    full_dl_l30[["CRAN"]],
    width = 3,
    icon = "fa fa-cloud-download",
    style = "info"
  ),
  summaryBox(
    "PyPI Downloads (Last 30 days)",
    full_dl_l30[["PyPI"]],
    width = 3,
    icon = "fa fa-cloud-download",
    style = "info"
  ),
  summaryBox(
    "CRAN Downloads (Lifetime)",
    cran_lifetime,
    width = 3,
    icon = "fa fa-trophy",
    style = "success"
  ),
  summaryBox(
    paste0("PyPI Downloads (Lifetime, ", pypi_lifetime_source, ")"),
    pypi_lifetime,
    width = 3,
    icon = "fa fa-trophy",
    style = "success"
  )
)

```


# Download statistics - CRAN, PyPI

The following plot shows the download statistics of the libraries from CRAN (R) and PyPI (Python). 

In general, download statistics from CRAN and PyPI should be more representative of actual user download and updates.

## Plots

### Daily downloads from CRAN and PyPI

Thin lines show raw daily counts; thicker lines show a 14-day centred rolling mean to highlight the trend without hiding individual spikes.

```{r daily-plot, echo=FALSE,message=FALSE,warning=FALSE}
roll_mean <- function(x, k = 14) {
  stats::filter(x, rep(1 / k, k), sides = 2) %>% as.numeric()
}

plot_data <-
  full_dl %>%
  arrange(date) %>%
  mutate(
    CRAN_smooth = roll_mean(CRAN),
    PyPI_smooth = roll_mean(PyPI)
  ) %>%
  select(date, CRAN, PyPI, CRAN_smooth, PyPI_smooth) %>%
  tidyr::pivot_longer(
    cols = -date,
    names_to = c("source", "series"),
    names_sep = "_",
    values_to = "downloads"
  ) %>%
  mutate(series = ifelse(is.na(series), "raw", series))

ts_plot <-
  ggplot(plot_data, aes(x = date, y = downloads, colour = source, alpha = series, linewidth = series)) +
  geom_line() +
  scale_colour_manual(values = c("CRAN" = "#1f77b4", "PyPI" = "#ff7f0e")) +
  scale_alpha_manual(values = c("raw" = 0.3, "smooth" = 1), guide = "none") +
  scale_linewidth_manual(values = c("raw" = 0.4, "smooth" = 1.1), guide = "none") +
  labs(title = "CRAN and PyPI Downloads",
       x = "Date",
       y = "Downloads",
       colour = "Source") +
  theme_minimal() +
  theme(plot.title = element_text(size = 16, face = "bold"),
        axis.title = element_text(size = 14),
        axis.text = element_text(size = 12),
        legend.title = element_text(size = 12),
        legend.text = element_text(size = 12))

plotly::ggplotly(ts_plot)
```

## Notes

- The download figures may sometimes show irregularities, as new releases tend to trigger more downloads due to automated package archiving.
- Package updates also count towards downloads